Delocalization of Eigenvectors in Random Matrices

### 2025-03-19 Draft.tex
fixed identity (3)

\subsection*{Random Band Matrices (RBM)}
The Random band matrix model is of great interest by itself because it serves as an interpolation between Wigner matrices and the random Schrödinger operator $H_{RS}$ \cite{dynamic}. A RBM $(H_{xy})_{x, y\in \Gamma}$, with centered entries independent up to symmetry $H_{ij} = \overline{H_{ji}}$, can represent a $d$-dimensional quantum system on a graph $\Gamma = \llbracket 1, N\rrbracket ^d$ with the effective distance being of order $W<\frac{N}{2}$, defined to the band width of the model \cite{7, knowles2013}.  What this means is that for $|i-j|_{N} = \min\left([x-y]_N, [y-x]_N\right)>W$ we have $H_{ij}=0$. It is standard to also normalize the covariance matrix $S_{xy} = \mathbb{E} |H_{xy}|^2$, s.t. $\sum_x S_{xy}=1$ for any $x\in \Gamma$. The model is conjectured to exhibit both localization for $W\ll W_c$ and delocalization for $W\gg W_c$ w.r.t a critical transition band $W_c$ that depends on the band width $W$ and the dimension $d$ \cite{7}:  \[
W_c = 
\begin{cases} 
\sqrt{N} & \text{for } d=1 \\ 
\sqrt{\log N} & \text{for } d=2 \\
O(1) & \text{for } d\geq 3

@@

The matrix model that will be subject to our examination throughout this paper is defined as follows: identify $\Z_n$ with $\Gamma=\{1, \cdots, N\}$ and equip it with the aforementioned periodic distance $|i-j|_N$. Now, for a symmetric and compactly supported probability density function (PDF) on $\mathbb{R}$, define the doubly stochastic matrix as: $$S_{xy} := Z_{N, W}^{-1}f\left(\frac{|x-y|_N}{W}\right),$$
where $Z_{N, W}\asymp W$ is a normalizing constant bounded deterministically above and below by $W$. A technical, but necessary assumption is for $S$ to admit a matrix square root $S^{1/2}$, satisfying the same properties for a similarly symmetric and compactly supported PDF $f'$. \vspace{0.5 cm}
\begin{center}
\begin{tikzpicture}
    % Define parameters
    \def\n{4} % Matrix dimension
    \def\w{0.5} % Half-width of the band

@@

    % Add zeros in the correct regions (outside the band)
    \node at (\n/4*3+0.5,\n/4*3-0.5) {$0$};
     \node at (\n/4*3-0.5,\n/4*3+0.5) {$0$};
    \node at (\n/4-0.5,\n/4+0.5) {$0$};
        \node at (\n/4+0.5,\n/4-0.5) {$0$};
\end{tikzpicture}
\end{center}

\newpage
\section{Prerequisites}
The main tool used to study both Anderson localization and random matrix statistics is the \textit{Green function} or \textit{resolvent}. For a random matrix $H$, it is defined as: $$G(z) := (H-zI_N)^{-1} = (H-z)^{-1},$$
where $z\in \mathbb{C}\backslash \{\lambda_1, \cdots, \lambda_N\}$ is the spectral parameter ranging over the complement of the eigenvalues $\lambda_1(H)\leq \cdots \leq \lambda_N(H)$ of $H$. Observe that for any eigenvector $\mathbf{u}_\alpha$ we have:
$$H\mathbf{u}_\alpha = \lambda_\alpha \mathbf{u}\Leftrightarrow (H - z)\mathbf{u}_\alpha = (\lambda_\alpha-z)\mathbf{u}_\alpha$$
\begin{equation}
\Rightarrow G(z)\mathbf{u}_\alpha = (H-z)^{-1}\mathbf{u}_\alpha = \frac{1}{\lambda_\alpha-z}\frac{(\lambda_\alpha-z)\mathbf{u}_\alpha }{H-z}=\frac{1}{\lambda_\alpha -z}\mathbf{u}_\alpha\end{equation}

\noindent This is equivalent to $\mathbf{u}_\alpha$ being eigenvectors for the resolvent with $\frac{1}{\lambda_\alpha-z}$ as the eigenvalues. This also means that we have by spectral decomposition 
$$G(z) =\sum_{i=1}^N \frac{\mathbf{u}_\alpha\mathbf{u}_\alpha^*}{\lambda_\alpha-z}.$$
Now, recall that their empirical measure for $H$ is defined as \cite{dynamic}: $$\varrho_N(dx):=\frac{1}{N}\sum_{j=1}^N \delta(x-\lambda_j)dx,$$
Let $m(z)=m_N(z) = \int_\R \frac{\varrho_N(x)}{x-z}dx$ be its Stieltjes transform for $z = E+i\eta$, where $E\in \R$ and $\eta>0$. By the relationship we found above, we have the following  identity, which follows from the definition of the Dirac delta function: $$m(z)= \int_\R \frac{\varrho_N(x)}{x-z}dx = \int_\mathbb{R} \frac{1}{N}\frac{\sum_{j=1}^N \delta(x-\lambda_j)dx}{x-z}dx = \frac{1}{N}\sum_{j=1}^N\int_\R\frac{\delta(x-\lambda_j)}{x-z}dx =\frac{1}{N} \sum_{j=1}^N \frac{1}{\lambda_j-z} = \frac{1}{N}\text{Tr}(G(z))$$
From the moments method \cite{dynamic} we know that the Stieltjes transform $m_N(z)$ of the empirical distribution converges in probability to the Stieltjes transform $m_{SC} = \frac{1}{2}(-z+\sqrt{z^2-4})$ of the semicircle law as $N\rightarrow\infty$. This means that in the limit, with probability one the self-consistent equation holds: \begin{equation}
m(z) = -\frac{1}{z+m(z)} \Leftrightarrow m(z) + \frac{1}{m(z)} + z=0
\end{equation}
Observe that: $$m(z)(m(z)+z) = \frac{1}{4}(-z+\sqrt{z^2-4})(z+\sqrt{z^2-4})=\frac{1}{4}(-z^2+z^2-4)=-1$$
$$\Rightarrow |m(z)| = \frac{1}{|m(z)+z|^{-1}}$$
Now, take the imaginary part of $(2)$: $$\text{Im}(m(z)) + \text{Im}\left(\frac{1}{m(z)}\right)+\eta=0=\text{Im}(m(z)) + \frac{\text{Im}}{|m(z)|^2}+\eta$$
$$\Rightarrow \text{Im}(m(z))\left(1-\frac{1}{|m(z)|^2}\right)=-\eta$$
And since $\text{Im}(m(z))>0$ and $\eta>0$, we must have $1-\frac{1}{|m(z)|^2}<0$, i.e $|m(z)|<1$. 
$$\Rightarrow \text{Im }m \asymp 1\quad 1-|m|^2 \asymp \eta$$

\subsection*{Resolvent Identities}
We can define the augmented minor Green function for any index $\alpha$ as $G(z)^{(\alpha)} = (H^{(\alpha)}-w_t)^{-1}$, where: $$\left(H^{(\alpha)}\right):=H_{ab}\cdot \mathbf{1}(a\neq \alpha)\cdot \mathbf{1}(b\neq \alpha),$$ 
so the augmented matrix is still $N\times N$ with the $\alpha$-row and -column set to zero \cite{dynamic}. As a consequence: 
\begin{equation}
G_{xy}^{(\alpha)} = \left\{\begin{array}{cc}
  (-z)^{-1}  & x=y=i\\
  0 & x=\alpha\ \text{XOR } y = \alpha\\
  G_{xy}^{[\alpha]} = (H^{[\alpha]}-z)_{xy}^{-1} & x\neq \alpha, y\neq \alpha
  \end{array}
  \right.
\end{equation}
Before we derive the estimate, let us first establish several identities:\newpage

>>>>

\subsection*{Random Band Matrices (RBM)}
The Random band matrix model is of great interest by itself because it serves as an interpolation between Wigner matrices and the random Schrödinger operator $H_{RS}$ \cite{dynamic}. A RBM $(H_{xy})_{x, y\in \Gamma}$, with centered complex random variables (r.v.), independent up to symmetry $H_{ij} = \overline{H_{ji}}$, can represent a $d$-dimensional quantum system on a graph $\Gamma = \llbracket 1, N\rrbracket ^d$ with the effective distance being of order $W<\frac{N}{2}$, defined to the band width of the model \cite{7, knowles2013}.  What this means is that for $|i-j|_{N} = \min\left([x-y]_N, [y-x]_N\right)>W$ we have $H_{ij}=0$. It is standard to also normalize the covariance matrix $S_{xy} = \mathbb{E} |H_{xy}|^2$, s.t. $\sum_x S_{xy}=1$ for any $x\in \Gamma$. The model is conjectured to exhibit both localization for $W\ll W_c$ and delocalization for $W\gg W_c$ w.r.t a critical transition band $W_c$ that depends on the band width $W$ and the dimension $d$ \cite{7}:  \[
W_c = 
\begin{cases} 
\sqrt{N} & \text{for } d=1 \\ 
\sqrt{\log N} & \text{for } d=2 \\
O(1) & \text{for } d\geq 3

@@

The matrix model $(H_{ij})_{i, j\in \Gamma}$ that will be subject to our examination throughout this paper is defined as follows: identify $\Z_n$ with $\Gamma=\{1, \cdots, N\}$ and equip it with the aforementioned periodic distance $|i-j|_N$, s.t. $\forall_{|i-j|_N>W} H_{ij}=0$. Now, for a symmetric and compactly supported probability density function (PDF) on $\mathbb{R}$, define the doubly stochastic matrix $(S_{xy})_{x, y\in \Gamma}$ as: $$\mathbb{E}|H_{xy}|^2=S_{xy} := Z_{N, W}^{-1}f\left(\frac{|x-y|_N}{W}\right),$$
where $Z_{N, W}\asymp W$ is a normalizing constant bounded deterministically above and below by $W$. A technical, but necessary assumption is for $S$ to admit a matrix square root $S^{1/2}$, satisfying the same properties for a similarly symmetric and compactly supported PDF $f'$. \vspace{0.75 cm}
\begin{center}
\begin{tikzpicture}
    % Define parameters
    \def\n{4} % Matrix dimension
    \def\w{0.5} % Half-width of the band

@@

    % Add zeros in the correct regions (outside the band)
    \node at (\n/4*3+0.5,\n/4*3-0.5) {$0$};
     \node at (\n/4*3-0.5,\n/4*3+0.5) {$0$};
    \node at (\n/4-0.5,\n/4+0.5) {$0$};
        \node at (\n/4+0.5,\n/4-0.5) {$0$};
\end{tikzpicture}
\end{center}
\vspace{0.25 cm}

\noindent In our analysis, we will utilize the flow method (\hyperref[sec:flow]{Section 3}), where the spectral parameter $z=E+i\eta$ of the resolvent of $H$ (defined in \hyperref[sec:prereq]{Section 2}) is varied at constant-speed in the upper-half plane $\mathbb{H}$ ($E\in\mathbb{R}$ and $\eta>0$) as the entries of $H$ are realized as Brownian motion \cite{bandSDE}.

\newpage
\section{Prerequisites}
\label{sec:prereq}
The main tool used to study both Anderson localization and random matrix statistics is the \textit{Green function} or \textit{resolvent}. For a random matrix $H$, it is defined as: $$G(z) := (H-zI_N)^{-1} = (H-z)^{-1},$$
where $z\in \mathbb{C}\backslash \{\lambda_1, \cdots, \lambda_N\}$ is the spectral parameter ranging over the complement of the eigenvalues $\lambda_1(H)\leq \cdots \leq \lambda_N(H)$ of $H$. Observe that for any eigenvector $\mathbf{u}_\alpha$ we have:
$$H\mathbf{u}_\alpha = \lambda_\alpha \mathbf{u}\Leftrightarrow (H - z)\mathbf{u}_\alpha = (\lambda_\alpha-z)\mathbf{u}_\alpha$$
\begin{equation}
\Rightarrow G(z)\mathbf{u}_\alpha = (H-z)^{-1}\mathbf{u}_\alpha = \frac{1}{\lambda_\alpha-z}\frac{(\lambda_\alpha-z)\mathbf{u}_\alpha }{H-z}=\frac{1}{\lambda_\alpha -z}\mathbf{u}_\alpha\end{equation}

\noindent This is equivalent to $\mathbf{u}_\alpha$ being eigenvectors for the resolvent with $\frac{1}{\lambda_\alpha-z}$ as the eigenvalues. This also means that we have by spectral decomposition 
$$G(z) =\sum_{i=1}^N \frac{\mathbf{u}_\alpha\mathbf{u}_\alpha^*}{\lambda_\alpha-z}.$$
Now, recall that their empirical measure for $H$ is defined as \cite{dynamic}: $$\varrho_N(dx):=\frac{1}{N}\sum_{j=1}^N \delta(x-\lambda_j)dx,$$
Let $m(z)=m_N(z) = \int_\R \frac{\varrho_N(x)}{x-z}dx$ be its Stieltjes transform for $z = E+i\eta$, where $E\in \R$ and $\eta>0$. By the relationship we found above, we have the following  identity, which follows from the definition of the Dirac delta function: $$m(z)= \int_\R \frac{\varrho_N(x)}{x-z}dx = \int_\mathbb{R} \frac{1}{N}\frac{\sum_{j=1}^N \delta(x-\lambda_j)dx}{x-z}dx = \frac{1}{N}\sum_{j=1}^N\int_\R\frac{\delta(x-\lambda_j)}{x-z}dx =\frac{1}{N} \sum_{j=1}^N \frac{1}{\lambda_j-z} = \frac{1}{N}\text{Tr}(G(z))$$
From the moments method \cite{dynamic} we know that the Stieltjes transform $m_N(z)$ of the empirical distribution converges in probability to the Stieltjes transform $m_{SC} = \frac{1}{2}(-z+\sqrt{z^2-4})$ of the semicircle law as $N\rightarrow\infty$. This means that in the limit, with probability one the self-consistent equation holds: \begin{equation}
m(z) = -\frac{1}{z+m(z)} \Leftrightarrow m(z) + \frac{1}{m(z)} + z=0
\end{equation}
Observe that: $$m(z)(m(z)+z) = \frac{1}{4}(-z+\sqrt{z^2-4})(z+\sqrt{z^2-4})=\frac{1}{4}(-z^2+z^2-4)=-1$$
$$\Rightarrow |m(z)| = \frac{1}{|m(z)+z|^{-1}}$$
Now, take the imaginary part of $(2)$: $$\text{Im}(m(z)) + \text{Im}\left(\frac{1}{m(z)}\right)+\eta=0$$
We can calculate the second term by simply expanding: $$\frac{1}{m(z)} = \frac{1}{\text{Re}(m(z))+\text{Im}(m(z))i} = \frac{\text{Re}(m(z))-\text{Im}(m(z))i}{|m(z)|^2}\Leftrightarrow \text{Im}\left(\frac{1}{m(z)}\right) = -\frac{\text{Im}(m(z))}{|m(z)|^2}$$
$$\Rightarrow\text{Im}(m(z)) + \text{Im}\left(\frac{1}{m(z)}\right)+\eta=\text{Im}(m(z)) -\frac{\text{Im}(m(z))}{|m(z)|^2}+\eta=0$$
$$\Rightarrow \text{Im}(m(z))\left(1-\frac{1}{|m(z)|^2}\right)=-\eta$$
And since $\text{Im}(m(z))>0$ and $\eta>0$, we must have $1-\frac{1}{|m(z)|^2}<0$, i.e $|m(z)|<1$, which is equivalent to $\text{Im }m \asymp 1$, where $\asymp$ means being bounded above and below up to a fixed, positive factor. Furthermore, let us rewrite the latter equation as the following identity: \begin{equation}1-|m(z)|^2 = \frac{\eta|m(z)|^2}{\text{Im}(m(z))}
\end{equation}

$$ 1-|m|^2 \asymp \eta$$

\subsection*{Resolvent Identities}
We can define the augmented minor Green function for any index $\alpha$ as $G(z)^{(\alpha)} = (H^{(\alpha)}-w_t)^{-1}$, where: $$\left(H^{(\alpha)}\right):=H_{ab}\cdot \mathbf{1}(a\neq \alpha)\cdot \mathbf{1}(b\neq \alpha),$$ 
so the augmented matrix is still $N\times N$ with the $\alpha$-row and -column set to zero \cite{dynamic}. As a consequence: 
\begin{equation}
G_{xy}^{(\alpha)} = \left\{\begin{array}{cc}
  (-z)^{-1}  & x=y=i\\
  0 & x=\alpha\ \text{XOR } y = \alpha\\
  G_{xy}^{[\alpha]} = (H^{[\alpha]}-z)_{xy}^{-1} & x\neq \alpha, y\neq \alpha
  \end{array}
  \right.
\end{equation}
Before we derive the estimate, let us first establish several identities:
